---
title: "OligoMet Profiler: A Standard End-to-End Workflow"
author:
- name: Nishikant Wase, PhD
  email: nishikant.wase@gmail.com
  affiliation: Research Scientist, Thermo Fisher Scientific (author and developer)
date: "`r Sys.Date()`"
package: OligoMetProfiler
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    number_sections: true
abstract: >
  A single worked example, start to finish: build a sequence, generate its
  theoretical metabolite library, import raw mzML files, obtain putative
  MS1 and MS2 identifications, and run a two-group statistical comparison.
  The example uses six small synthetic mzML files bundled with the package
  (three "control" and three "treated" replicates of the inotersen
  reference sequence, plus a known-ground-truth JSON file used to build
  them), so every chunk here runs on its own with no external data and no
  Python installation. For the full option set -- notations, the chemistry
  dictionary, custom overrides, PS oxidation, acquisition-method export,
  spectral libraries, and the parallel batch pipeline for many-sample
  studies -- see `vignette("OligoMetProfiler")`.
vignette: >
  %\VignetteIndexEntry{OligoMet Profiler: A Standard End-to-End Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  FOR RESEARCH USE ONLY: not for diagnostic, clinical, or
  regulatory submission use. All outputs are computed
  predictions, not measurements, and must be confirmed
  experimentally; the software is provided without warranty
  and the author accepts no liability for its use.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

Metabolite identification for a therapeutic oligonucleotide is a targeted
problem: a finite set of plausible biotransformation products (exonuclease
truncations, endonuclease fragments, PS-to-PO oxidation states), each with
a calculable mass, checked against what actually eluted in an LC-MS run.
This vignette runs that whole chain once, on one drug, so the shape of the
pipeline is visible before the option tables in the main vignette. Four
stages, each its own section below:

1. **Sequence** -- write the drug's chemistry as a triplet string, parse it.
2. **Library** -- generate the candidate metabolites (parent plus a
   truncation ladder) from that sequence.
3. **Raw data** -- read six mzML files (a small synthetic run standing in
   for a real acquisition) and pull out MS1 features and MS2 spectra.
4. **Identification and statistics** -- match the library against each
   file's MS1 features, confirm hits against MS2 fragments, then compare
   intensities between two experimental groups.

```{r library}
library(OligoMetProfiler)
```

# The example dataset

Six mzML files ship under `inst/extdata/batch_example/`: three replicates
labeled "control" and three labeled "treated", built by
`inst/extdata/batch_example/generate_example.py` from a fixed random seed.
Each file carries a 7-scan chromatographic peak for the inotersen parent
across four charge states, the same 7-scan peak for two truncated
degradants (a 3' N-1 and a 5' N-1, at roughly a third of the parent's
level), a contaminant trace at an unrelated m/z that matches nothing in
the library, and one MS2 scan holding real McLuckey fragment ions plus
the two phosphorothioate diagnostic ions. The treated replicates were
generated at several-fold higher intensity than control for all three
real species, so the comparison in the final section has a genuine,
known-direction effect to recover. `theoretical_values.json` in the same
folder records the exact masses used to build the files, for anyone who
wants to check the pipeline's output against ground truth directly.

```{r locate-data}
example_dir <- system.file("extdata", "batch_example", package = "OligoMetProfiler")
files <- list.files(example_dir, pattern = "\\.mzML$", full.names = TRUE)
sample_meta <- read.csv(file.path(example_dir, "sample_meta.csv"),
                        stringsAsFactors = FALSE)
sample_meta
```

# Building the sequence

Inotersen is a 20-mer, 5-10-5 MOE gapmer: MOE wings at positions 1-5 and
16-20, a DNA gap at 6-15, phosphorothioate throughout, five
5-methylcytosine positions. Written as a triplet string -- one
`[linkage][base][sugar]` token per position, 5' to 3' -- that sequence is
the `INOTERSEN_TRIPLET` constant shipped with the package:

```{r build-sequence}
spec <- parse_input(INOTERSEN_TRIPLET)
format_spec(spec)
```

`parse_input()` also reads two other notations (the `OH-...-OH` form used
by OligoDistiller, and a three-line bases/sugars/linkages layout closer to
what a Certificate of Analysis or a BioPharma Finder export gives you),
plus a fully structured R list for anything a triplet string cannot
express, such as terminal conjugates. Those, along with how to add a
sugar or backbone code the built-in dictionary lacks, are covered in
`vignette("OligoMetProfiler")`.

# Building the metabolite library

`generate_metabolites()` turns a parsed spec into the candidate list: the
parent plus, in this run, a 3-deep truncation ladder from each terminus
(`max_3p`/`max_5p`). Endonuclease fragments are switched off here
(`endo = FALSE`) since this particular dataset carries only the parent and
the two shortmers -- a real run of a gapmer would usually enable them.

```{r build-library}
mets <- generate_metabolites(spec, opts = list(
  oligo_name = "inotersen", max_3p = 3, max_5p = 3, endo = FALSE))
length(mets)
metabolite_table(mets)[, c("id", "name", "kind", "n")]
```

Seven candidates: the parent, three 3'-truncations (N-1 through N-3), and
three 5'-truncations. Only two of those six truncations are actually
present in the example data -- the library is a hypothesis space, not a
claim that every candidate was observed, and the matching step below
sorts out which ones were.

# Importing raw mzML files

`read_ms_file()` reads one mzML (or mzXML) file into MS1 peaks, MS2
spectra, and file metadata. It prefers Bioconductor's `Spectra`/`mzR` when
both are installed and falls back to a built-in xml2-based parser
otherwise -- the same result either way, so nothing below depends on which
path is taken. Vendor raw formats (Thermo `.raw`, Sciex `.wiff`,
Bruker/Agilent) go through `resolve_ms_input_file()` first, which converts
via ProteoWizard `msconvert` when it is on the PATH.

```{r read-one-file}
raw <- read_ms_file(files[1])
raw$info
head(raw$ms1)
```

`raw$ms1` is a flat peak table (`rt`, `mz`, `intensity`); `raw$ms2` adds
`precursor_mz` and `precursor_z` per row, since one file can hold many
MS2 scans on different precursors. Reading raw MS1 peaks is not the same
as having features to match against -- consecutive scans of the same
eluting ion need to be collapsed into one entry first, which is what the
next section does.

# MS1 output: putative identification

For each file, `extract_ms1_features()` groups raw peaks that co-elute in
time and agree in m/z into single features, and `match_ms1()` checks
every (metabolite, PS-oxidation level, charge state, adduct) combination
from the library against those features within a ppm window. Charge
states here are restricted to 5-8 and the adduct list to plain `[M-zH]`
because that is what this particular synthetic acquisition covers; an
unknown real sample should start from the wider defaults shown in the
main vignette (`z_range = 3:12`, `adducts = c("H", "Na", "K", "NH4")`).

```{r ms1-matching}
ms1_matches <- do.call(rbind, lapply(seq_along(files), function(i) {
  raw  <- read_ms_file(files[i])
  feat <- extract_ms1_features(raw$ms1)
  hit  <- match_ms1(mets, feat, ppm_tol = 10, z_range = 5:8,
                     adducts = "H", max_oxid = 0, n_iso = 0)
  if (nrow(hit) > 0) hit$sample <- sample_meta$sample[i]
  hit
}))

unique(ms1_matches[, c("met_id", "met_name", "kind")])
```

Three of the seven candidates come back with a match -- the parent and
one truncation from each end -- across all six files, at whichever
charge states and ppm accuracy each file's noise happened to produce:

```{r ms1-table}
head(ms1_matches[, c("sample", "met_name", "z", "theo_mz",
                     "obs_mz", "ppm_error", "intensity")], 8)
```

This loop is doing by hand what `match_ms1_batch()`/
`annotate_metabolites_batch()` do automatically once a multi-sample
feature table exists -- those functions are the right entry point for a
real study of more than a handful of files, and are covered together with
the parallel Python deconvolution pipeline in the main vignette's batch
section.

# MS2 output: putative identification

A single MS1 match is a mass coincidence until something independent
confirms it. `find_ms2_spectra()` locates the MS/MS scan recorded against
a given precursor m/z and charge, and `confirm_metabolite()` scores it:
McLuckey fragment coverage, number of matched fragment ions, mass
accuracy, and the presence of the two phosphorothioate diagnostic ions
(m/z 94.9452 and 192.9746), combined into a 0-100 confirmation score.
Inotersen's fully phosphorothioate backbone means those diagnostic ions
should turn up in every file's MS2 scan.

```{r ms2-confirmation}
parent_info  <- metabolite_mass_info(mets[[1]])
parent_z6_mz <- charge_envelope(parent_info$mono_mass, z_range = 6)$mz

ms2_confirm <- do.call(rbind, lapply(seq_along(files), function(i) {
  raw  <- read_ms_file(files[i])
  spec <- find_ms2_spectra(raw$ms2, precursor_mz = parent_z6_mz,
                           precursor_z = 6, ppm_tol = 20)
  if (length(spec) == 0) return(NULL)
  conf <- confirm_metabolite(mets[[1]], spec[[1]], tol_ppm = 15, z_range = 1:2)
  data.frame(sample = sample_meta$sample[i],
             n_ms2_peaks = nrow(spec[[1]]),
             n_frag_matches = conf$score$n_matches,
             coverage = round(conf$score$coverage, 2),
             n_diagnostics = conf$score$n_diagnostics,
             confirmation_score = conf$score$total_score,
             confident = conf$score$confident)
}))
ms2_confirm
```

`confident` is `TRUE` once the total score clears 75 and sequence coverage
clears 50%, a threshold following the FMVS confirmation-scoring approach
cited in the main vignette. Shown here for the parent only, at one charge
state, for a single metabolite at a time; `confirm_ms2_batch()` runs the
same confirmation across every MS1 hit in a combined multi-sample table.

# Statistical analysis: control versus treated

`build_abundance_matrix()` collapses the combined MS1 matches into one
row per metabolite, one column per sample, taking the most intense match
across charge states as that metabolite's abundance in that sample.
`abundance_long()` reshapes it against the sample metadata, and
`compare_two_groups()` runs a Welch t-test per metabolite with
Benjamini-Hochberg-adjusted p-values.

```{r abundance-matrix}
abund <- build_abundance_matrix(ms1_matches)
abund
```

```{r statistics}
long   <- abundance_long(abund, sample_meta)
result <- compare_two_groups(long, "control", "treated")
result[order(result$p_adj),
       c("met_id", "met_name", "mean_a", "mean_b", "log2fc", "p_value", "p_adj")]
```

All three species were generated at higher intensity under "treated" --
roughly threefold for the parent and a bit over twofold for the two
degradants -- so a positive `log2fc` with a small `p_adj` on every row is
the expected recovery of that built-in design, not a coincidence of this
particular random seed. `plot_volcano()` renders the same comparison as a
fold-change-versus-significance plot:

```{r volcano, fig.width = 5, fig.height = 4}
plot_volcano(result, fc_thresh = 0.5)
```

`compare_multi_groups()` (one-way ANOVA plus Tukey HSD) and
`compare_time_series()` (linear trend) cover three-or-more-group and
timepoint designs the same way; both are documented alongside this
function in the main vignette.

# Where to go from here

This walkthrough stops at the point where a real study usually branches:

* **Wider chemistry** -- other input notations, the chemistry dictionary,
  custom overrides for a code the dictionary lacks, and the PS oxidation
  series -- `vignette("OligoMetProfiler")`.
* **Many-sample studies** -- the bundled Python pipeline
  (`run_batch_deconvolution()`) does chromatographic feature detection
  across a whole directory of raw files at once, in place of the
  per-file loop shown above; `run_batch_ms.R` wraps it end to end with
  the same statistics functions used here.
* **Deliverables** -- `build_workbook()` for the multi-sheet Excel
  library, `build_report()` for an HTML/PDF summary,
  `thermo_ms1_inclusion_list()`/`thermo_ms2_prm_target_list()` for
  Orbitrap Exploris acquisition method export, and
  `export_spectral_libraries()` for MGF/MSP spectral libraries.
* **A dashboard** -- `run_app()` wraps the same engine in a Shiny UI with
  file upload, an editable custom-chemistry table, and downloadable
  outputs.

# Session info {.unnumbered}

```{r sessioninfo}
sessionInfo()
```

# Disclaimer {.unnumbered}

**Research use only.** Every metabolite, mass, match, and statistic in
this vignette is a computed prediction from synthetic example data, not a
measurement from a real sample, and nothing here is validated for
diagnostic, clinical, or regulatory use. See `DISCLAIMER.md` and the
disclosure section at the end of `vignette("OligoMetProfiler")` for the
full statement, authorship, and licensing terms.
