---
title: "OligoMet Profiler: In-Silico Metabolite Identification for Therapeutic Oligonucleotides"
author:
- name: Nishikant Wase, PhD
  email: nishikant.wase@gmail.com
  affiliation: Research Scientist, Thermo Fisher Scientific (author and developer)
date: "`r Sys.Date()`"
package: OligoMetProfiler
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
abstract: >
  OligoMetProfiler generates theoretical metabolite libraries, molecular
  formulas, negative-ESI charge envelopes, isotope patterns,
  phosphorothioate (PS) oxidation series, McLuckey MS/MS fragment ions,
  and PRM inclusion lists for any therapeutic oligonucleotide -- DNA,
  RNA, 2'-O-methyl, 2'-fluoro, MOE, cEt, LNA, mixed PO/PS backbones, and
  fully custom chemistry via dictionary overrides. It optionally matches
  the theoretical library against acquired MS data (mzML/mzXML or peak
  lists), exports a seven-sheet Excel workbook and an HTML/PDF report,
  and writes targeted mass lists formatted for the Thermo Orbitrap
  Exploris Method Editor. This vignette walks through the complete
  workflow: sequence input in three notations, the chemistry dictionary
  and custom overrides, metabolite library generation, mass and envelope
  calculation, fragment ion prediction, acquisition method export, MS
  data matching, and the Shiny dashboard.
vignette: >
  %\VignetteIndexEntry{OligoMet Profiler: In-Silico Metabolite Identification for Therapeutic Oligonucleotides}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  FOR RESEARCH USE ONLY: not for diagnostic, clinical, or
  regulatory submission use. All outputs are computed
  predictions, not measurements, and must be confirmed
  experimentally; the software is provided without warranty
  and the author accepts no liability for its use.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Introduction

Metabolite identification for therapeutic oligonucleotides (antisense
oligonucleotides, siRNA, and related modalities) requires predicting the
masses of every plausible biotransformation product before any of them
can be found in LC-MS data. Exonucleases chew nucleotides from the 3'
and 5' termini, endonucleases cleave internal linkages,
phosphorothioate (PS) backbones progressively desulfurize to
phosphodiester (PO) *in vivo* and *in vitro*, and each of these species
ionizes as a multi-charge negative-ESI envelope with sodium, potassium,
and ammonium adduct satellites. Doing this by hand in a spreadsheet is
error-prone and does not scale past a handful of metabolites.

`OligoMetProfiler` automates the whole chain. Given a sequence and its
chemistry (sugars, linkages, terminal conjugates), it produces:

* **Theoretical metabolite library** -- parent, 3'/5' exonuclease
  truncation ladders, and endonuclease internal fragments.
* **Masses and formulas** -- elemental formula, monoisotopic mass, and
  average mass for every metabolite, assembled from a validated
  chemistry dictionary.
* **Charge envelopes** -- $[M - zH]^{z-}$ m/z values across a
  configurable charge-state range, with H/Na/K/NH~4~ adduct variants.
* **Isotope patterns** -- top-N isotope peaks per charge state, via
  `enviPat` when installed or a built-in convolution engine otherwise.
* **PS oxidation series** -- the -15.977 Da PS-to-PO desulfurization
  ladder at each oxidation level.
* **McLuckey MS/MS fragment ions** -- a, a-B, b, b-B, c, d, w, x, y, z
  terminal ions plus internal fragments, with PS diagnostic ions,
  following the nomenclature of McLuckey and co-workers.
* **Acquisition support** -- a PRM inclusion list, plus Orbitrap
  Exploris Method Editor-formatted MS1 inclusion and MS2 PRM target
  lists.
* **Reporting** -- a seven-sheet Excel workbook and an HTML/PDF report
  with summary plots.
* **MS matching (optional)** -- import mzML/mzXML or peak lists, match
  MS1 features against the library with ppm tolerance, score isotope
  fit and envelope consistency, and confirm hits via MS/MS fragment
  matching.

The methodology is grounded in four published approaches: SynONIM
(Lippens *et al.*, JASMS 2024) for synthetic oligonucleotide
modification and impurity nomenclature, the eluforsen metabolite
profiling study (Kim *et al.*, Mol Ther Nucleic Acids 2019) for the
LC-MS/MS workflow and PS diagnostic ions, OligoDistiller (Liu *et
al.*, Anal Chem 2025) for McLuckey fragment annotation and isotope
envelope fitting, and the FMVS automatic metabolite identification
method (Ye *et al.*, J Chromatogr B 2025) for confirmation scoring.

Four approved oligonucleotide therapeutics ship with the package as
worked examples, one per modality class in Takakusa *et al.* (2023):
**nusinersen** (antisense SSO), **inotersen** (antisense gapmer),
**patisiran** (siRNA delivered by lipid nanoparticle) and **givosiran**
(GalNAc-conjugated siRNA). The two antisense examples double as the
formula engine's regression anchors -- the pipeline reproduces their
published free-acid molecular formulas exactly. They are illustrations,
not default targets: every function in this vignette runs on any
sequence you give it.

# Installation

Install directly from GitHub:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("nishi76/OligoMet-Profiler")
```

The hard dependencies (`openxlsx`, `ggplot2`, `xml2`, `xfun`) are
installed automatically. Two optional packages improve specific steps:

* `enviPat` -- higher-accuracy isotope patterns (a built-in convolution
  engine is used when it is absent);
* `rmarkdown` -- required only for rendering HTML/PDF reports.

The Shiny dashboard additionally uses `shiny`, `DT`, `bslib`, and
`shinyFiles`; those are pulled in by `install_packages.R` in the
repository root, or install them manually.

```{r library}
library(OligoMetProfiler)
```

# Quick start

A complete run on a 16-mer gapmer ASO -- 2'-O-methyl wings, DNA gap,
all-PS backbone -- takes four calls:

```{r quickstart}
seq_str <- "Gm-sTm-sCm-sTm-sCm-sTd-sCd-sTd-sCd-sTd-sTd-sCm-sTm-sCm-sTm-sGm"

spec <- parse_input(seq_str)
mets <- generate_metabolites(spec, opts = list(
  oligo_name = "demo", max_3p = 5, max_5p = 5, endo = TRUE))

info <- metabolite_mass_info(mets[[1]])
cat("Parent formula:", info$formula_str,
    "\nMonoisotopic mass:", round(info$mono_mass, 4), "Da\n")

length(mets)   # parent + truncations + endonuclease fragments
```

The Excel workbook and HTML report wrap everything downstream of this
into single calls (`build_workbook()`, `build_report()`); they are
covered in the *Outputs* section below.

# From analytical documentation to an input string

The pipeline needs three pieces of information *per nucleotide
position*, plus anything attached to the ends. All of it is normally
on the documentation that accompanies the oligonucleotide -- a
Certificate of Analysis (CoA), a synthesis report, a patent sequence
listing, or a BioPharma Finder sequence-editor export:

1. **Base sequence** (5' to 3'). Watch for 5-methylcytosine: CoAs
   write it as `mC`, `5-Me-C`, `C*`, or a footnote such as "all
   cytosines are 5-methyl". Its base code here is `S`, not `C` -- the
   two differ by CH~2~ (14.016 Da) per site.
2. **Sugar at each position**, often given as a design shorthand:
   "5-10-5 MOE gapmer" means 5 MOE positions, 10 DNA, 5 MOE;
   "2'-OMe/2'-F alternating" alternates `m` and `f`.
3. **Backbone linkages** -- "full phosphorothioate" is the common case
   for ASOs; mixed backbones are listed per position or as a pattern
   ("PS wings, PO gap"). An *n*-mer has *n - 1* linkages.
4. **Terminal conjugates** (GalNAc, cholesterol, amino linkers,
   phosphate caps, dyes) -- absent unless stated.

To construct the triplet string, write one dash-separated token per
position, 5' to 3', as `[incoming linkage][BASE][sugar]`, leaving the
linkage prefix off the first token.

Take inotersen as a worked case. Its published description reads:

> 20-mer antisense oligonucleotide targeting the TTR 3'UTR (bases
> 618-637), sequence 5'-TCTTGGTTACATGAAATCCC-3'. A 5-10-5 gapmer:
> positions 1-5 and 16-20 are 2'-MOE, positions 6-15 are 2'-deoxy.
> Fully phosphorothioate backbone. All cytosines are 5-methylcytosine.

Translating position by position:

| Position | Base (from CoA) | Base code | Sugar | Incoming linkage | Token |
|---:|---|---|---|---|---|
| 1 | T | `T` | 2'-MOE -> `e` | (none, 5' end) | `Te` |
| 2 | 5-Me-C | `S` | 2'-MOE -> `e` | PS -> `s` | `sSe` |
| 6 | G | `G` | DNA -> `d` (gap starts) | PS -> `s` | `sGd` |
| 16 | A | `A` | 2'-MOE -> `e` (wing resumes) | PS -> `s` | `sAe` |
| 20 | 5-Me-C | `S` | 2'-MOE -> `e` | PS -> `s` | `sSe` |

Joined up, that is the `INOTERSEN_TRIPLET` constant shipped with the
package:

```{r coa-example}
spec_coa <- parse_input(INOTERSEN_TRIPLET)
spec_coa$n
paste(spec_coa$bases, collapse = "")
paste(spec_coa$sugars, collapse = "")
```

The proof that the translation is right is the molecular formula: the
assembled formula matches the one on the approved product labelling
exactly.

```{r coa-check}
info_coa <- metabolite_mass_info(list(
  bases = spec_coa$bases, sugars = spec_coa$sugars,
  linkages = spec_coa$linkages, conj5 = "none", conj3 = "none"))
info_coa$formula_str
INOTERSEN_FORMULA            # published free-acid formula
round(info_coa$avg_mass, 2)  # published average MW: 7183.08
```

Two traps are worth calling out, because both change the mass and
neither is obvious from the letter sequence alone. Writing `C` instead
of `S` for a 5-methylcytosine loses CH~2~ (14.016 Da) *per site* -- in
inotersen that is five sites, a 70 Da error. And an *n*-mer has
*n - 1* linkages, so the last token never carries an outgoing bond.

The parser enforces the obvious sanity checks -- token count equals
oligo length, every token after the first starts with a linkage code,
and any code missing from the dictionary is named in the error
message. Modifications without a built-in code are added as custom
overrides with their elemental formula (see *Custom chemistry
overrides* below). The package's `inst/help/QUICKSTART.md`
carries a condensed version of this guide with the full code tables.

## The four bundled reference drugs

One approved therapeutic per modality class ships with the package,
covering the chemistries you are most likely to meet:

```{r reference-table}
do.call(rbind, lapply(REFERENCE_OLIGOS, function(r) data.frame(
  drug = r$name, brand = r$brand, modality = r$modality,
  nt = r$length, chemistry = r$chemistry, stringsAsFactors = FALSE)))
```

Each is available as a triplet string (`NUSINERSEN_TRIPLET`,
`INOTERSEN_TRIPLET`, `PATISIRAN_SENSE_TRIPLET`,
`GIVOSIRAN_SENSE_TRIPLET`) and through the `REFERENCE_OLIGOS` registry,
which also carries the terminal conjugates that triplet notation cannot
express:

```{r reference-run}
for (r in REFERENCE_OLIGOS) {
  sp <- parse_input(r$triplet)
  info <- metabolite_mass_info(list(
    bases = sp$bases, sugars = sp$sugars, linkages = sp$linkages,
    conj5 = r$conj5, conj3 = r$conj3))
  cat(sprintf("%-26s %-20s %2d nt  %-38s %10.4f Da\n",
              r$name, r$modality, sp$n, info$formula_str, info$mono_mass))
}
```

Note what distinguishes each modality, since these are exactly the
features that drive metabolite prediction:

* **nusinersen** -- uniform 2'-MOE and a fully PS backbone. Sugar-modified
  throughout, so it degrades by exonuclease nibbling from the termini
  rather than internal cleavage.
* **inotersen** -- a 5-10-5 gapmer, MOE wings around a DNA gap. The gap is
  the endonuclease-susceptible region, so `endo_sites = "gap"` matches
  its real biotransformation route.
* **patisiran** (sense strand) -- an all-phosphodiester siRNA with a dTdT
  overhang. No PS means no oxidation series to model (`max_oxid` has
  nothing to act on).
* **givosiran** (sense strand) -- alternating 2'-OMe/2'-F, PS only at the
  two 5'-terminal linkages, and a trivalent GalNAc at the 3' end. The
  conjugate rides on the parent and on 5' truncations, but is lost from
  3' truncations that cut past it.

Duplex drugs are given as their sense strand: this pipeline profiles one
strand at a time, so run each strand of an siRNA separately.

# Sequence input notations

`parse_input()` auto-detects three input formats. All three reduce to
the same internal representation: parallel vectors of bases, sugars,
and linkages (5' to 3'), plus optional 5'/3' conjugate codes.

## Triplet notation

Each dash-separated token is `[linkage][base][sugar]`. The first (5')
token has no linkage prefix; the prefix on token *i* is the incoming
bond from position *i - 1*.

```{r triplet}
spec <- parse_input(
  "Gm-sTm-sCm-sTm-sCm-sTd-sCd-sTd-sCd-sTd-sTd-sCm-sTm-sCm-sTm-sGm")
spec$n
paste(spec$bases, collapse = "")
paste(spec$sugars, collapse = "")
```

Breaking the example down:

* `Gm` -- position 1: base G, sugar `m` (2'-OMe), no incoming linkage.
* `sTm` -- position 2: base T, sugar `m`, PS bond from position 1.
* `sTd` -- position 6: base T, sugar `d` (DNA), PS bond from position 5.

Linkage codes are `o` or `p` for phosphodiester and `s` for
phosphorothioate. `p` matches Thermo BioPharma Finder's sequence-editor
notation, so triplet strings copied from BPF (e.g. `Ad-pTd-pCd-pAd`)
parse directly.

## OligoDistiller notation

Wrapped in `OH-...-OH`, with `*` marking a PS *outgoing* bond and the
sugar as a suffix on the base:

```{r oligodistiller}
spec_od <- parse_input(
  "OH-Gm*-Tm*-Cm*-Tm*-Cm*-Td*-Cd*-Td*-Cd*-Td*-Td*-Cm*-Tm*-Cm*-Tm*-Gm-OH")
identical(spec_od$bases, spec$bases)
```

Sugar suffixes: `m` = 2'-OMe, `f` = 2'-F, `d` = DNA, `r` = RNA,
`e` = MOE.

## Three-line notation (bases / sugars / linkages)

The layout a chemical analysis file or a BioPharma Finder sequence entry
gives you directly, and the one the dashboard's **Manual sequence entry**
panel uses. One code per position for bases and sugars; one per *bond*
for linkages, so an n-mer has n-1 of them.

```{r three-line}
spec_3l <- parse_three_line(
  bases    = "TSASTTTSATAATGSTGG",
  sugars   = "eeeeeeeeeeeeeeeeee",
  linkages = "sssssssssssssssss")
format_triplet(spec_3l)
```

A field holding a single code is applied to every position, which is the
common case for a uniformly modified drug -- `sugars = "e"` and
`linkages = "s"` give exactly the result above. Multi-character codes
(`MOE`, `cEt`, `LNA`) need a separator: `"MOE-MOE-d-d"`, `"A,G,S,T"`.

`format_three_line()` renders any spec back to the three lines, and
`format_biopharma_fasta()` writes it as a FASTA record for BioPharma
Finder's Sequence Manager:

```{r bpf-fasta}
cat(format_biopharma_fasta(spec_3l, name = "nusinersen"))
```

See `inst/help/SEQUENCE_GUIDE.md` (or the dashboard's Help panel) for a
walkthrough aimed at readers who have
not worked with oligonucleotide notation before.

## Structured input

An R list with explicit parallel vectors -- use this when you need
terminal conjugates or full programmatic control. `linkages[i]` is the
bond from position *i* to *i + 1*; the last entry is `NA` (no outgoing
bond at the 3' terminus).

```{r structured}
my_spec <- parse_input(list(
  bases    = c("A", "C", "G", "U", "A", "G", "U", "C", "A", "C", "G", "U"),
  sugars   = c("m", "f", "m", "f", "m", "f", "m", "f", "m", "f", "m", "f"),
  linkages = c("s", "s", "s", "s", "s", "s", "s", "s", "s", "s", "s", NA),
  conj5    = "none",
  conj3    = "GalNAc"
))
my_spec$conj3
```

`validate_spec()` checks any parsed spec against the dictionary and
reports unknown codes before you get halfway through a run.

# The chemistry dictionary

Every module resolves base, sugar, linkage, and conjugate codes through
a chemistry dictionary. `STANDARD_DICT` is the built-in default:

```{r dict-peek}
# A couple of representative entries
STANDARD_DICT[["m"]]$name
STANDARD_DICT[["s"]]$name
format_formula(STANDARD_DICT[["G"]]$formula)
```

## Built-in codes

Bases: `A` (adenine), `G` (guanine), `C` (cytosine), `T` (thymine),
`U` (uracil), `S` (5-methylcytosine), `D` (2,6-diaminopurine),
`I` (hypoxanthine/inosine).

Sugars: `d` (2'-deoxyribose), `r` (ribose), `m` (2'-O-methylribose),
`f` (2'-fluororibose), `FANA` (2'-F arabino; isobaric with `f`),
`e`/`MOE` (2'-O-methoxyethyl), `NMA` (2'-O-[2-(methylamino)-2-oxoethyl]),
`allyl` (2'-O-allyl), `AP` (2'-O-aminopropyl), `NH2`
(2'-amino-2'-deoxyribose), `UNA` (unlocked, 2',3'-seco), `GNA` (glycol
nucleic acid), `LNA`, `ENA` (2'-O,4'-C-ethylene), `cEt` (constrained
ethyl), `NP` (3'-amino, for N3'->P5' phosphoramidate backbones).

Linkages: `o`/`p` (phosphodiester), `s` (phosphorothioate), `u`
(PS stereochemistry variant), the alkyl phosphonates `mp` (methyl),
`prp` (propyl), `ibu` (isobutyl), `chx` (cyclohexyl) and `mop`
(methoxypropyl), `pace`/`tpace` (phosphonoacetate and its thio
analog), `msp` (mesyl phosphoramidate), and `pgo`/`tmg` (the two
phosphoryl guanidine variants -- they differ by 2.0157 Da per bond, so
check which one you have).

Conjugates: 5'/3'-phosphate caps, GalNAc (mono and trivalent cluster),
cholesterol, C6/C12 amino linkers, TEG, FAM, Cy3, the fatty acids
`myristoyl`/`palmitoyl`/`stearoyl`/`docosanoyl`, plus the BioPharma
Finder 5.2 terminal-modification set (biotin, cAG/cAU/ARCA/mCAP cap
analogs, triantennary GalNAc).

`inst/help/MODIFICATIONS.md` -- the Modifications tab of the
dashboard's Help panel -- lists every one of these with its residue
formula and its per-position mass difference from the parent chemistry,
and explains how to add a modification the dictionary does not have.

Entries flagged `verify = TRUE` (e.g. `cEt`, `LNA`, `ENA`, `msp`,
`pgo`, `GalNAc3`, dye conjugates) carry best-estimate formulas --
confirm them against a known mass before relying on them, or override
them (next section).

## Custom chemistry overrides

Any code can be added or replaced with `build_dictionary()`. Each
override is a named list carrying a `formula` (named vector or string)
and an optional `name`; conjugates also take an `attach` mode
(`replace_H`, `replace_OH`, or `add`).

```{r overrides}
custom_dict <- build_dictionary(list(
  # A new base code
  X = list(formula = c(C = 6, H = 7, N = 5, O = 1),
           name = "my modified base X"),
  # A new sugar code, formula as string
  Q = list(formula = "C10H20O7", name = "my 2'-mod sugar Q"),
  # Replace a best-estimate entry with a confirmed formula
  cEt = list(formula = "C7H12O4", name = "constrained ethyl (confirmed)")
))
format_formula(custom_dict[["Q"]]$formula)
```

Pass `dict = custom_dict` to any downstream function, and use the new
codes in sequences exactly like standard ones (`Gm-sXm-sQm-...`). The
pipeline trusts custom formulas as given -- there is no external
validation gate -- so when a reference mass for the full-length oligo
is available, compare it against the computed parent mass to catch
transcription errors.

## The formula-engine self-test

`validate_reference()` re-derives the molecular formulas of nusinersen
and inotersen from their sequences and chemistry, and compares each
against the free-acid formula on its approved product labelling.
Between them they exercise MOE and DNA sugars, 5-methylcytosine and
5-methyluracil bases, and a fully phosphorothioate backbone. It is a
regression check on the formula engine (run it after editing the
dictionary), not a validation of whatever sequence you are analyzing:

```{r ion337}
v <- validate_reference(verbose = FALSE)
v$ok                       # TRUE only if both formulas reproduce exactly
vapply(v$all, function(x) x$formula, "")
vapply(v$all, function(x) round(x$ppm, 6), 0)   # deviation, ppm
```

# Metabolite library generation

`generate_metabolites()` builds the library from a parsed spec:

```{r metabolites}
mets <- generate_metabolites(spec, opts = list(
  oligo_name   = "demo",
  max_3p       = 5,      # 3' exonuclease ladder depth
  max_5p       = 5,      # 5' exonuclease ladder depth
  endo         = TRUE,   # endonuclease internal fragments
  endo_sites   = "all",  # or "gap" to cleave only in the DNA gap
  min_frag_len = 3       # discard fragments shorter than this
))
head(metabolite_table(mets), 8)
```

Options worth knowing:

* **`max_3p` / `max_5p`** -- ladder depth. For an *n*-mer, `n - 1`
  gives the complete series; the default (10) covers the metabolites
  actually observed in most in vivo studies.
* **`endo_sites = "gap"`** -- for gapmer ASOs, endonucleases (S1,
  RNase H-adjacent activities) preferentially cleave within the DNA
  gap. Restricting cleavage to the gap keeps the library focused and
  the downstream envelope computation fast.
* **`min_frag_len`** -- short fragments produce low-m/z, low-charge
  species that rarely survive LC-MS detection; 3 is a practical floor.

Each metabolite records its own bases/sugars/linkages (so conjugate
loss on truncation is handled correctly), its kind (`parent`,
`exo_3p`, `exo_5p`, `endo_5p`/`endo_3p`), PS/PO linkage counts, and a
human-readable modification label.

# Masses, envelopes, isotopes, and the PS oxidation series

## Mass and formula

```{r massinfo}
info <- metabolite_mass_info(mets[[1]])
info$formula_str
round(info$mono_mass, 6)
round(info$avg_mass, 4)
```

The formula is assembled as: sum of bases + sum of sugars + sum of
internal linkages - (2n - 1) H~2~O + conjugates, accounting for *n*
glycosidic bonds and *n - 1* internucleotide bonds. This convention is
validated against the published molecular formulas of nusinersen and
inotersen (0 ppm on both).

## Charge envelope

Negative ESI, $[M - zH]^{z-}$:

```{r envelope}
charge_envelope(info$mono_mass, z_range = 4:8)
```

`h_offset` shifts the neutral mass before charging. Leave it at 0 for
standard mass spectrometry; a non-zero value exists only to reproduce
legacy reference workbooks that computed their envelopes from an
offset neutral mass.

Adduct satellites replace one H with a cation; the shifts are exposed
via `adduct_shift()`:

```{r adducts}
sapply(c("Na", "K", "NH4"), adduct_shift)
```

## Isotope patterns

`isotope_pattern()` returns the top-N isotopologues for a formula.
With `enviPat` installed it uses enviPat's convolution; otherwise a
built-in engine takes over (identical selection logic, slightly
coarser peak merging). Both paths are memoized per formula, so
repeated calls across charge states are effectively free.

```{r isotopes}
pat <- isotope_pattern(info$formula_vec, n_top = 5, use_envipat = FALSE)
pat$mass <- round(pat$mass, 4)
pat
```

`isotope_mz_cluster()` converts a pattern into per-charge-state m/z
values (the monoisotopic peak is always retained even when it is not
among the most abundant), and `compute_envelope()` produces the full
tidy table -- one row per (oxidation level, charge state, isotope) --
that feeds the workbook's Charge Envelopes sheet:

```{r compute-envelope}
env <- compute_envelope(mets[[1]], z_range = 5:6, n_iso = 3,
                        max_oxid = 1, use_envipat = FALSE)
env$mz <- round(env$mz, 4)
head(env[, c("met_id", "k_oxid", "z", "iso", "mz", "abundance")], 6)
```

## PS oxidation series

Each PS-to-PO desulfurization replaces one S with one O
(-15.977 Da). `ps_oxidation_series()` walks the ladder from 0 to
`min(n_PS, max_oxid)` events:

```{r oxidation}
ser <- ps_oxidation_series(mets[[1]], max_oxid = 3)
ser$mono_mass <- round(ser$mono_mass, 4)
ser[, c("k", "mono_mass", "formula_str")]
```

Heavily oxidized species are low-abundance in practice; the default
cap of 6 events covers what is typically observable.

# MS/MS fragment ions

`generate_fragments()` produces McLuckey terminal ions -- a, a-B, b,
b-B, c, d from the 5' side and w, x, y, z from the 3' side -- and
`generate_internal_fragments()` adds internal (double-cleavage)
products:

```{r fragments}
frags <- generate_fragments(mets[[1]])
length(frags)
ft <- fragment_table(frags)
head(ft[ft$z == 1, c("ion_type", "direction", "cleavage_site",
                     "frag_length", "mz")], 8)
```

For PS backbones, `check_ps_diagnostic()` looks for the low-mass PS
diagnostic ions (94.9362 / 192.9660 series) in an MS2 spectrum, and
`confirmation_score()` combines fragment coverage, diagnostic ions,
and mass accuracy into a 0-100 confirmation score per metabolite
(following the FMVS approach).

# Outputs: workbook, report, acquisition lists

## Excel workbook

`build_workbook()` writes the seven-sheet library workbook: Summary,
Metabolite Library, Charge Envelopes, PS Oxidation Series, Fragment
Ions, PRM Inclusion List, and MS Matching.

```{r workbook, eval = FALSE}
opts <- list(z_range = 3:12, n_iso = 5, max_oxid = 6, h_offset = 0,
             use_envipat = TRUE, include_internal = TRUE,
             max_3p = 10, max_5p = 10, endo = TRUE,
             adducts = c("H", "Na", "K", "NH4"), ppm_tol = 10)

wb_path <- build_workbook(spec, mets, opts = opts,
                          output_file = "demo_library.xlsx",
                          output_dir  = ".")
```

The Charge Envelopes sheet is the most computation-heavy (one isotope
pattern per metabolite per oxidation level); a live console progress
bar with an adaptive ETA tracks it. Sheets are assembled in memory and
written in bulk, so the Excel serialization itself takes seconds even
for libraries with tens of thousands of envelope rows.

## HTML/PDF report

`build_report()` renders a standalone report (requires `rmarkdown`)
with five summary plots -- charge envelope, isotope pattern,
truncation series, oxidation series, and fragment map. The individual
`plot_*()` functions are exported too, so any plot can be regenerated
or restyled on its own:

```{r report, eval = FALSE}
report <- build_report(spec, mets, opts = opts,
                       output_file = "demo_report",
                       output_format = "html",   # or "pdf"
                       plot_dir = ".")
```

## Orbitrap Exploris acquisition method export

Three targeted-acquisition tables are exported in the exact column
layout of the Orbitrap Exploris Method Editor's Targeted Mass filter
(Mass List Type "m/z & z", Time Mode "Start/End Time"):

```{r acquisition}
ms1 <- thermo_ms1_inclusion_list(mets[1:3], z_range = 5:8,
                                 max_oxid = 1, max_targets = 500)
head(ms1, 4)
```

* **`thermo_ms1_inclusion_list()`** -- every (metabolite, oxidation
  level, charge state) combination; import into a Full Scan
  experiment's Targeted Mass filter.
* **`thermo_ms2_prm_target_list()`** -- the same list narrowed to a
  smaller charge range and tagged with an HCD collision energy; import
  as the target list of a Targeted MS2 (PRM) scan. The default NCE
  (20%) is a starting point for PS oligonucleotide backbones, not a
  validated parameter -- optimize per method.
* **`ms2_fragment_reference()`** -- theoretical fragment ions per
  metabolite. This is *not* a Method Editor import (PRM records the
  full fragment spectrum); it is the lookup table for interpreting the
  acquired spectra afterward.

Both list functions cap output rows (`max_targets`) -- the Method
Editor accepts up to 150,000 rows, but PRM duty cycle degrades long
before that.

## Spectral libraries (MGF / MSP)

The same theoretical library also exports as MGF and MSP, the formats
spectral-library tools read (MS-DIAL, mzVault/Compound Discoverer,
MZmine, matchms, SIRIUS, GNPS).

```{r spectral, eval = FALSE}
export_spectral_libraries(mets, dict, out_dir = "results",
                          prefix = "my_oligo")
#> results/my_oligo_MS1_library.mgf  (+ .msp)
#> results/my_oligo_MS2_library.mgf  (+ .msp)
```

`build_ms1_library()` emits one spectrum per (metabolite, PS-oxidation
level, charge state), whose peaks are the theoretical isotope cluster --
real relative abundances, annotated with the true isotopologue offset
(`M+0`, `M+1`, ...) and anchored on the m/z computed from the formula
rather than on the isotope engine's binned masses.

`build_ms2_library()` emits one spectrum per (metabolite, precursor
charge state), holding the McLuckey fragment ions at the requested
fragment charges, annotated `w4^2-`, `a-B5^1-`, `w-a(3,8)^1-` in the
MSP.

There is no fragment-intensity model in this pipeline, so **every MS2
peak is written at a flat intensity of 100**. Match on m/z; an
intensity-weighted dot-product or cosine score against these libraries
reduces to a function of peak count and means nothing. The MS1
libraries carry genuine isotope abundances and are not affected.

# Matching against acquired MS data

Point the pipeline at an mzML/mzXML file (or a two-column m/z-intensity
peak list) to annotate the library against real data:

```{r matching, eval = FALSE}
ms   <- parse_mzml("data.mzML")            # MS1 + MS2 spectra
feat <- extract_ms1_features(ms$ms1)
hits <- match_ms1(feat, mets, ppm_tol = 10,
                  z_range = 3:12, max_oxid = 6,
                  adducts = c("H", "Na", "K", "NH4"))
ann  <- annotate_metabolites(hits, ms, mets)   # adds MS2 confirmation
```

Matching proceeds in the order established by the eluforsen and FMVS
studies: (1) match each metabolite's theoretical m/z values across
charge states, adducts, and oxidation levels at the given ppm
tolerance; (2) score isotope-pattern fit (cosine similarity) and
charge-envelope consistency; (3) for hits with associated MS2 spectra,
confirm via McLuckey fragment matching and PS diagnostic ions.

Vendor raw files (Thermo `.raw`, Waters `.raw`, SCIEX `.wiff`) are
converted automatically when ProteoWizard `msconvert` is on the PATH.
mzML binary decoding uses a small Python helper when `python3` is
available and falls back to a pure-R decoder otherwise.

# The Shiny dashboard and CLI drivers

Three ready-made entry points wrap the same engine. The dashboard
ships inside the installed package; the CLI drivers live in the
repository root:

* **`run_app()`** -- a Shiny dashboard with sequence input (triplet, or
  the three-field manual entry described above), an editable
  custom-chemistry table, full parameter control, optional MS upload,
  a four-plot summary, and workbook/report/PRM/acquisition-list/
  spectral-library downloads. Launch with `OligoMetProfiler::run_app()` from an
  installed package, or `shiny::runApp(".")` from a clone of the
  repository.
* **`run_custom_oligo.R`** -- the primary CLI template: copy it, edit
  the CONFIG block (sequence, overrides, parameters), and
  `Rscript run_custom_oligo.R`.

For a formula-engine regression check from the command line, run the
validation scripts in `tests/` (e.g.
`Rscript tests/test_mass_isotope.R`) or call `validate_reference()`
directly from R.

# Performance notes

* The **Charge Envelopes** computation dominates runtime: one isotope
  pattern per metabolite per PS-oxidation level. Isotope patterns are
  memoized per formula (both the enviPat and built-in engines), so
  charge states and adducts add no repeated cost.
* To speed up a long run: disable endonuclease fragments, lower
  `max_oxid`, or narrow `z_range` -- each directly reduces the number
  of isotope-pattern calls.
* Workbook sheets are written in bulk (one write per sheet, styling
  restricted to headers and a few accent cells), so `saveWorkbook()`
  stays fast regardless of library size.
* `enviPat` prints a harmless note about the electron mass on some
  systems; it is cosmetic. Set `use_envipat = FALSE` to use the
  built-in engine instead.

# Caveats

* Dictionary entries flagged `verify = TRUE` (cEt, LNA, GalNAc3, dye
  and BPF terminal modifications) are best-estimate formulas --
  confirm against a known mass before quantitative use.
* Custom overrides are trusted as given; validate the parent mass
  against a reference when one exists.
* The `h_offset = 3.0046` convention exists only to reproduce legacy
  reference workbooks that used an offset neutral mass; use 0 for
  standard work.
* The repository's `tests/` scripts are console-output validation
  scripts, not asserted unit tests.

# Session info {.unnumbered}

```{r sessioninfo}
sessionInfo()
```

# References {.unnumbered}

## Methodology {.unnumbered}

1. Lippens J.L. *et al.* SynONIM: a Synthetic Oligonucleotide
   Nomenclature for Impurities and Modifications. *J Am Soc Mass
   Spectrom* (2024).
2. Kim J. *et al.* Metabolite Profiling of the Antisense
   Oligonucleotide Eluforsen Using Liquid Chromatography-Mass
   Spectrometry. *Mol Ther Nucleic Acids* (2019).
3. Liu R. *et al.* OligoDistiller: untargeted profiling of therapeutic
   oligonucleotide drugs and metabolites. *Anal Chem* (2025).
4. Ye X. *et al.* Automatic metabolite identification of antisense
   oligonucleotides by full MS variable scanning. *J Chromatogr B*
   (2025).
5. McLuckey S.A., Van Berkel G.J., Glish G.L. Tandem mass spectrometry
   of small, multiply charged oligonucleotides. *J Am Soc Mass
   Spectrom* 3:60-70 (1992).

## Reference drugs and modality classification {.unnumbered}

6. Takakusa H., Iwazaki N., Nishikawa M., Yoshida T., Obika S., Inoue T.
   Drug Metabolism and Pharmacokinetics of Antisense Oligonucleotide
   Therapeutics: Typical Profiles, Evaluation Approaches, and Points to
   Consider Compared with Small Molecule Drugs. *Nucleic Acid
   Therapeutics* 33(2):83-94 (2023). doi:10.1089/nat.2022.0054 --
   source of the modality classification (Table 1) used for the four
   bundled examples, and of the class-specific metabolism routes
   (endonuclease-first for gapmers, exonuclease for SSOs).
7. Egli M., Manoharan M. Chemistry, structure and function of approved
   oligonucleotide therapeutics. *Nucleic Acids Research*
   51(6):2529-2573 (2023). doi:10.1093/nar/gkad067 -- chemistry and
   structural review of the approved oligonucleotide drugs.
8. SPINRAZA (nusinersen) prescribing information, Biogen -- sequence,
   2'-MOE/PS chemistry, and free-acid molecular formula
   C~234~H~340~N~61~O~128~P~17~S~17~.
9. TEGSEDI (inotersen) prescribing information, Akcea/Ionis -- 5-10-5
   MOE gapmer design, TTR 3'UTR target sequence (bases 618-637), and
   free-acid molecular formula C~230~H~318~N~69~O~121~P~19~S~19~
   (average MW 7183.08).
10. ONPATTRO (patisiran) prescribing information, Alnylam -- siRNA
    duplex sequences, 2'-OMe modification pattern, dTdT overhangs, and
    lipid-nanoparticle formulation.
11. GIVLAARI (givosiran) prescribing information, Alnylam -- ESC-GalNAc
    siRNA sequences, 2'-F/2'-OMe alternation, terminal phosphorothioate
    placement, and the trivalent GalNAc (L96) 3' conjugate.

## Instrument and software documentation {.unnumbered}

12. Thermo Fisher Scientific. *BioPharma Finder 5.2 Oligonucleotide
    Analysis User Guide* -- triplet sequence notation and terminal
    modification codes.
13. Thermo Fisher Scientific. *Orbitrap Exploris 120 Software Manual*,
    "Targeted Inclusion -- Targeted Mass filter" -- mass-list column
    layout used by the acquisition exports.

# Author, disclosure and disclaimer

**Nishikant Wase, PhD** -- author and developer; Research Scientist,
Thermo Fisher Scientific (nishikant.wase@gmail.com). Designed, written
and maintained by the author; released under the MIT licence.

**Research use only.** OligoMetProfiler is a research tool. It is not a
medical device and is not intended or validated for diagnostic use,
clinical decision-making, patient care, quality control release testing,
or inclusion in a regulatory submission.

**Everything it reports is a prediction.** Libraries, formulas, masses,
envelopes, isotope patterns, fragment ions, target lists and spectral
libraries are computed from a chemistry dictionary and from assumptions
about oligonucleotide metabolism and fragmentation -- not measured. A
mass match is a hypothesis, not an identification; confirm every
assignment experimentally. Formulas flagged `verify = TRUE` are
unverified best estimates, and the MS2 spectral libraries carry
placeholder intensities.

**No warranty, no liability.** The software is provided "as is", without
warranty of any kind, express or implied, under the MIT licence. In no
event shall the author be liable for any claim, damages or other
liability -- including loss of data, wasted instrument time or
materials, or an erroneous scientific conclusion -- arising from or in
connection with the software or its use. The user is solely responsible
for verifying fitness for their purpose and for validating every result.

**Affiliation and disclosure.** The author is employed as a Research
Scientist at Thermo Fisher Scientific, disclosed here because the
package interoperates with Thermo Fisher products (BioPharma Finder
notation, Orbitrap Exploris mass lists). OligoMetProfiler is an
independent personal project, not a Thermo Fisher Scientific product,
and has not been supplied, reviewed, supported, endorsed or approved by
Thermo Fisher Scientific or any other company; all interoperability is
implemented from publicly documented formats only. Product and drug
names are used for identification and interoperability only and remain
the trademarks of their owners. Views and outputs are the author's own.

```{r about, eval = FALSE}
oligomet_about()   # prints this statement in the console
```

The full statement is in `DISCLAIMER.md`.
