---
title: "OligoMet Profiler: In-Silico Metabolite Identification for Therapeutic Oligonucleotides"
author:
- name: Nishikant Wase
  email: nishikant.wase@gmail.com
date: "`r Sys.Date()`"
package: OligoMetProfiler
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
abstract: >
  OligoMetProfiler generates theoretical metabolite libraries, molecular
  formulas, negative-ESI charge envelopes, isotope patterns,
  phosphorothioate (PS) oxidation series, McLuckey MS/MS fragment ions,
  and PRM inclusion lists for any therapeutic oligonucleotide -- DNA,
  RNA, 2'-O-methyl, 2'-fluoro, MOE, cEt, LNA, mixed PO/PS backbones, and
  fully custom chemistry via dictionary overrides. It optionally matches
  the theoretical library against acquired MS data (mzML/mzXML or peak
  lists), exports a seven-sheet Excel workbook and an HTML/PDF report,
  and writes targeted mass lists formatted for the Thermo Orbitrap
  Exploris Method Editor. This vignette walks through the complete
  workflow: sequence input in three notations, the chemistry dictionary
  and custom overrides, metabolite library generation, mass and envelope
  calculation, fragment ion prediction, acquisition method export, MS
  data matching, and the Shiny dashboard.
vignette: >
  %\VignetteIndexEntry{OligoMet Profiler: In-Silico Metabolite Identification for Therapeutic Oligonucleotides}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Introduction

Metabolite identification for therapeutic oligonucleotides (antisense
oligonucleotides, siRNA, and related modalities) requires predicting the
masses of every plausible biotransformation product before any of them
can be found in LC-MS data. Exonucleases chew nucleotides from the 3'
and 5' termini, endonucleases cleave internal linkages,
phosphorothioate (PS) backbones progressively desulfurize to
phosphodiester (PO) *in vivo* and *in vitro*, and each of these species
ionizes as a multi-charge negative-ESI envelope with sodium, potassium,
and ammonium adduct satellites. Doing this by hand in a spreadsheet is
error-prone and does not scale past a handful of metabolites.

`OligoMetProfiler` automates the whole chain. Given a sequence and its
chemistry (sugars, linkages, terminal conjugates), it produces:

* **Theoretical metabolite library** -- parent, 3'/5' exonuclease
  truncation ladders, and endonuclease internal fragments.
* **Masses and formulas** -- elemental formula, monoisotopic mass, and
  average mass for every metabolite, assembled from a validated
  chemistry dictionary.
* **Charge envelopes** -- $[M - zH]^{z-}$ m/z values across a
  configurable charge-state range, with H/Na/K/NH~4~ adduct variants.
* **Isotope patterns** -- top-N isotope peaks per charge state, via
  `enviPat` when installed or a built-in convolution engine otherwise.
* **PS oxidation series** -- the -15.977 Da PS-to-PO desulfurization
  ladder at each oxidation level.
* **McLuckey MS/MS fragment ions** -- a, a-B, b, b-B, c, d, w, x, y, z
  terminal ions plus internal fragments, with PS diagnostic ions,
  following the nomenclature of McLuckey and co-workers.
* **Acquisition support** -- a PRM inclusion list, plus Orbitrap
  Exploris Method Editor-formatted MS1 inclusion and MS2 PRM target
  lists.
* **Reporting** -- a seven-sheet Excel workbook and an HTML/PDF report
  with summary plots.
* **MS matching (optional)** -- import mzML/mzXML or peak lists, match
  MS1 features against the library with ppm tolerance, score isotope
  fit and envelope consistency, and confirm hits via MS/MS fragment
  matching.

The methodology is grounded in four published approaches: SynONIM
(Lippens *et al.*, JASMS 2024) for synthetic oligonucleotide
modification and impurity nomenclature, the eluforsen metabolite
profiling study (Kim *et al.*, Mol Ther Nucleic Acids 2019) for the
LC-MS/MS workflow and PS diagnostic ions, OligoDistiller (Liu *et
al.*, Anal Chem 2025) for McLuckey fragment annotation and isotope
envelope fitting, and the FMVS automatic metabolite identification
method (Ye *et al.*, J Chromatogr B 2025) for confirmation scoring.

The published gapmer ASO **ION337** ships with the package as a worked,
known-mass reference (validated to < 1 ppm against its literature
formula) -- it is a self-test for the formula engine, not a default
target. Every function in this vignette runs on any sequence you give
it.

# Installation

Install directly from GitHub:

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("nishi76/OligoMet-Profiler")
```

The hard dependencies (`openxlsx`, `ggplot2`, `xml2`, `xfun`) are
installed automatically. Two optional packages improve specific steps:

* `enviPat` -- higher-accuracy isotope patterns (a built-in convolution
  engine is used when it is absent);
* `rmarkdown` -- required only for rendering HTML/PDF reports.

The Shiny dashboard additionally uses `shiny`, `DT`, `bslib`, and
`shinyFiles`; those are pulled in by `install_packages.R` in the
repository root, or install them manually.

```{r library}
library(OligoMetProfiler)
```

# Quick start

A complete run on a 16-mer gapmer ASO -- 2'-O-methyl wings, DNA gap,
all-PS backbone -- takes four calls:

```{r quickstart}
seq_str <- "Gm-sTm-sCm-sTm-sCm-sTd-sCd-sTd-sCd-sTd-sTd-sCm-sTm-sCm-sTm-sGm"

spec <- parse_input(seq_str)
mets <- generate_metabolites(spec, opts = list(
  oligo_name = "demo", max_3p = 5, max_5p = 5, endo = TRUE))

info <- metabolite_mass_info(mets[[1]])
cat("Parent formula:", info$formula_str,
    "\nMonoisotopic mass:", round(info$mono_mass, 4), "Da\n")

length(mets)   # parent + truncations + endonuclease fragments
```

The Excel workbook and HTML report wrap everything downstream of this
into single calls (`build_workbook()`, `build_report()`); they are
covered in the *Outputs* section below.

# From analytical documentation to an input string

The pipeline needs three pieces of information *per nucleotide
position*, plus anything attached to the ends. All of it is normally
on the documentation that accompanies the oligonucleotide -- a
Certificate of Analysis (CoA), a synthesis report, a patent sequence
listing, or a BioPharma Finder sequence-editor export:

1. **Base sequence** (5' to 3'). Watch for 5-methylcytosine: CoAs
   write it as `mC`, `5-Me-C`, `C*`, or a footnote such as "all
   cytosines are 5-methyl". Its base code here is `S`, not `C` -- the
   two differ by CH~2~ (14.016 Da) per site.
2. **Sugar at each position**, often given as a design shorthand:
   "5-10-5 MOE gapmer" means 5 MOE positions, 10 DNA, 5 MOE;
   "2'-OMe/2'-F alternating" alternates `m` and `f`.
3. **Backbone linkages** -- "full phosphorothioate" is the common case
   for ASOs; mixed backbones are listed per position or as a pattern
   ("PS wings, PO gap"). An *n*-mer has *n - 1* linkages.
4. **Terminal conjugates** (GalNAc, cholesterol, amino linkers,
   phosphate caps, dyes) -- absent unless stated.

To construct the triplet string, write one dash-separated token per
position, 5' to 3', as `[incoming linkage][BASE][sugar]`, leaving the
linkage prefix off the first token. For a CoA that reads

> 16-mer ASO, sequence 5'-GTCTCTCTCTTCTCTG-3', 5-6-5 gapmer
> (positions 1-5 and 12-16 2'-O-methyl, 6-11 2'-deoxy), full
> phosphorothioate backbone, no conjugates

the translation is: position 1 = `Gm` (no prefix), position 2 =
`sTm` (PS bond in, base T, 2'-OMe), position 6 = `sTd` (DNA sugar
starts the gap), ..., position 16 = `sGm`:

```{r coa-example}
coa_seq <- "Gm-sTm-sCm-sTm-sCm-sTd-sCd-sTd-sCd-sTd-sTd-sCm-sTm-sCm-sTm-sGm"
spec_coa <- parse_input(coa_seq)
spec_coa$n
```

The parser enforces the obvious sanity checks -- token count equals
oligo length, every token after the first starts with a linkage code,
and any code missing from the dictionary is named in the error
message. Modifications without a built-in code are added as custom
overrides with their elemental formula (see *Custom chemistry
overrides* below). The repository's `QUICKSTART.md`
carries a condensed version of this guide with the full code tables.

# Sequence input notations

`parse_input()` auto-detects three input formats. All three reduce to
the same internal representation: parallel vectors of bases, sugars,
and linkages (5' to 3'), plus optional 5'/3' conjugate codes.

## Triplet notation

Each dash-separated token is `[linkage][base][sugar]`. The first (5')
token has no linkage prefix; the prefix on token *i* is the incoming
bond from position *i - 1*.

```{r triplet}
spec <- parse_input(
  "Gm-sTm-sCm-sTm-sCm-sTd-sCd-sTd-sCd-sTd-sTd-sCm-sTm-sCm-sTm-sGm")
spec$n
paste(spec$bases, collapse = "")
paste(spec$sugars, collapse = "")
```

Breaking the example down:

* `Gm` -- position 1: base G, sugar `m` (2'-OMe), no incoming linkage.
* `sTm` -- position 2: base T, sugar `m`, PS bond from position 1.
* `sTd` -- position 6: base T, sugar `d` (DNA), PS bond from position 5.

Linkage codes are `o` or `p` for phosphodiester and `s` for
phosphorothioate. `p` matches Thermo BioPharma Finder's sequence-editor
notation, so triplet strings copied from BPF (e.g. `Ad-pTd-pCd-pAd`)
parse directly.

## OligoDistiller notation

Wrapped in `OH-...-OH`, with `*` marking a PS *outgoing* bond and the
sugar as a suffix on the base:

```{r oligodistiller}
spec_od <- parse_input(
  "OH-Gm*-Tm*-Cm*-Tm*-Cm*-Td*-Cd*-Td*-Cd*-Td*-Td*-Cm*-Tm*-Cm*-Tm*-Gm-OH")
identical(spec_od$bases, spec$bases)
```

Sugar suffixes: `m` = 2'-OMe, `f` = 2'-F, `d` = DNA, `r` = RNA,
`e` = MOE.

## Structured input

An R list with explicit parallel vectors -- use this when you need
terminal conjugates or full programmatic control. `linkages[i]` is the
bond from position *i* to *i + 1*; the last entry is `NA` (no outgoing
bond at the 3' terminus).

```{r structured}
my_spec <- parse_input(list(
  bases    = c("A", "C", "G", "U", "A", "G", "U", "C", "A", "C", "G", "U"),
  sugars   = c("m", "f", "m", "f", "m", "f", "m", "f", "m", "f", "m", "f"),
  linkages = c("s", "s", "s", "s", "s", "s", "s", "s", "s", "s", "s", NA),
  conj5    = "none",
  conj3    = "GalNAc"
))
my_spec$conj3
```

`validate_spec()` checks any parsed spec against the dictionary and
reports unknown codes before you get halfway through a run.

# The chemistry dictionary

Every module resolves base, sugar, linkage, and conjugate codes through
a chemistry dictionary. `STANDARD_DICT` is the built-in default:

```{r dict-peek}
# A couple of representative entries
STANDARD_DICT[["m"]]$name
STANDARD_DICT[["s"]]$name
format_formula(STANDARD_DICT[["G"]]$formula)
```

## Built-in codes

Bases: `A` (adenine), `G` (guanine), `C` (cytosine), `T` (thymine),
`U` (uracil), `S` (5-methylcytosine), `D` (2,6-diaminopurine),
`I` (hypoxanthine/inosine).

Sugars: `d` (2'-deoxyribose), `r` (ribose), `m` (2'-O-methylribose),
`f` (2'-fluororibose), `e`/`MOE` (2'-O-methoxyethylribose), `NH2`
(2'-amino-2'-deoxyribose), `cEt` (constrained ethyl), `LNA` (locked
nucleic acid).

Linkages: `o`/`p` (phosphodiester), `s` (phosphorothioate), `u`
(PS stereochemistry variant).

Conjugates: 5'/3'-phosphate caps, GalNAc (mono and trivalent cluster),
cholesterol, C6/C12 amino linkers, TEG, FAM, Cy3, plus the BioPharma
Finder 5.2 terminal-modification set (biotin, cAG/cAU/ARCA/mCAP cap
analogs, triantennary GalNAc).

Entries flagged `verify = TRUE` (e.g. `cEt`, `LNA`, `GalNAc3`, dye
conjugates) carry best-estimate formulas -- confirm them against a
known mass before relying on them, or override them (next section).

## Custom chemistry overrides

Any code can be added or replaced with `build_dictionary()`. Each
override is a named list carrying a `formula` (named vector or string)
and an optional `name`; conjugates also take an `attach` mode
(`replace_H`, `replace_OH`, or `add`).

```{r overrides}
custom_dict <- build_dictionary(list(
  # A new base code
  X = list(formula = c(C = 6, H = 7, N = 5, O = 1),
           name = "my modified base X"),
  # A new sugar code, formula as string
  Q = list(formula = "C10H20O7", name = "my 2'-mod sugar Q"),
  # Replace a best-estimate entry with a confirmed formula
  cEt = list(formula = "C7H12O4", name = "constrained ethyl (confirmed)")
))
format_formula(custom_dict[["Q"]]$formula)
```

Pass `dict = custom_dict` to any downstream function, and use the new
codes in sequences exactly like standard ones (`Gm-sXm-sQm-...`). The
pipeline trusts custom formulas as given -- there is no external
validation gate -- so when a reference mass for the full-length oligo
is available, compare it against the computed parent mass to catch
transcription errors.

## The ION337 reference self-test

`validate_ION337()` re-derives the published ION337 formula and mass
from its triplet string and reports the deviation. It is a regression
check on the formula engine (run it after editing the dictionary), not
a validation of whatever sequence you are analyzing:

```{r ion337}
v <- validate_ION337(verbose = FALSE)
v$formula
v$ok
round(v$ppm, 4)   # deviation from the literature mass, ppm
```

# Metabolite library generation

`generate_metabolites()` builds the library from a parsed spec:

```{r metabolites}
mets <- generate_metabolites(spec, opts = list(
  oligo_name   = "demo",
  max_3p       = 5,      # 3' exonuclease ladder depth
  max_5p       = 5,      # 5' exonuclease ladder depth
  endo         = TRUE,   # endonuclease internal fragments
  endo_sites   = "all",  # or "gap" to cleave only in the DNA gap
  min_frag_len = 3       # discard fragments shorter than this
))
head(metabolite_table(mets), 8)
```

Options worth knowing:

* **`max_3p` / `max_5p`** -- ladder depth. For an *n*-mer, `n - 1`
  gives the complete series; the default (10) covers the metabolites
  actually observed in most in vivo studies.
* **`endo_sites = "gap"`** -- for gapmer ASOs, endonucleases (S1,
  RNase H-adjacent activities) preferentially cleave within the DNA
  gap. Restricting cleavage to the gap keeps the library focused and
  the downstream envelope computation fast.
* **`min_frag_len`** -- short fragments produce low-m/z, low-charge
  species that rarely survive LC-MS detection; 3 is a practical floor.

Each metabolite records its own bases/sugars/linkages (so conjugate
loss on truncation is handled correctly), its kind (`parent`,
`exo_3p`, `exo_5p`, `endo_5p`/`endo_3p`), PS/PO linkage counts, and a
human-readable modification label.

# Masses, envelopes, isotopes, and the PS oxidation series

## Mass and formula

```{r massinfo}
info <- metabolite_mass_info(mets[[1]])
info$formula_str
round(info$mono_mass, 6)
round(info$avg_mass, 4)
```

The formula is assembled as: sum of bases + sum of sugars + sum of
internal linkages - (2n - 1) H~2~O + conjugates, accounting for *n*
glycosidic bonds and *n - 1* internucleotide bonds. This convention is
validated to < 1 ppm against the ION337 literature mass.

## Charge envelope

Negative ESI, $[M - zH]^{z-}$:

```{r envelope}
charge_envelope(info$mono_mass, z_range = 4:8)
```

`h_offset` shifts the neutral mass before charging. Leave it at 0 for
standard mass spectrometry; `3.0046` reproduces the empirical
convention used by the ION337 reference workbook.

Adduct satellites replace one H with a cation; the shifts are exposed
via `adduct_shift()`:

```{r adducts}
sapply(c("Na", "K", "NH4"), adduct_shift)
```

## Isotope patterns

`isotope_pattern()` returns the top-N isotopologues for a formula.
With `enviPat` installed it uses enviPat's convolution; otherwise a
built-in engine takes over (identical selection logic, slightly
coarser peak merging). Both paths are memoized per formula, so
repeated calls across charge states are effectively free.

```{r isotopes}
pat <- isotope_pattern(info$formula_vec, n_top = 5, use_envipat = FALSE)
pat$mass <- round(pat$mass, 4)
pat
```

`isotope_mz_cluster()` converts a pattern into per-charge-state m/z
values (the monoisotopic peak is always retained even when it is not
among the most abundant), and `compute_envelope()` produces the full
tidy table -- one row per (oxidation level, charge state, isotope) --
that feeds the workbook's Charge Envelopes sheet:

```{r compute-envelope}
env <- compute_envelope(mets[[1]], z_range = 5:6, n_iso = 3,
                        max_oxid = 1, use_envipat = FALSE)
env$mz <- round(env$mz, 4)
head(env[, c("met_id", "k_oxid", "z", "iso", "mz", "abundance")], 6)
```

## PS oxidation series

Each PS-to-PO desulfurization replaces one S with one O
(-15.977 Da). `ps_oxidation_series()` walks the ladder from 0 to
`min(n_PS, max_oxid)` events:

```{r oxidation}
ser <- ps_oxidation_series(mets[[1]], max_oxid = 3)
ser$mono_mass <- round(ser$mono_mass, 4)
ser[, c("k", "mono_mass", "formula_str")]
```

Heavily oxidized species are low-abundance in practice; the default
cap of 6 events covers what is typically observable.

# MS/MS fragment ions

`generate_fragments()` produces McLuckey terminal ions -- a, a-B, b,
b-B, c, d from the 5' side and w, x, y, z from the 3' side -- and
`generate_internal_fragments()` adds internal (double-cleavage)
products:

```{r fragments}
frags <- generate_fragments(mets[[1]])
length(frags)
ft <- fragment_table(frags)
head(ft[ft$z == 1, c("ion_type", "direction", "cleavage_site",
                     "frag_length", "mz")], 8)
```

For PS backbones, `check_ps_diagnostic()` looks for the low-mass PS
diagnostic ions (94.9362 / 192.9660 series) in an MS2 spectrum, and
`confirmation_score()` combines fragment coverage, diagnostic ions,
and mass accuracy into a 0-100 confirmation score per metabolite
(following the FMVS approach).

# Outputs: workbook, report, acquisition lists

## Excel workbook

`build_workbook()` writes the seven-sheet library workbook: Summary,
Metabolite Library, Charge Envelopes, PS Oxidation Series, Fragment
Ions, PRM Inclusion List, and MS Matching.

```{r workbook, eval = FALSE}
opts <- list(z_range = 3:12, n_iso = 5, max_oxid = 6, h_offset = 0,
             use_envipat = TRUE, include_internal = TRUE,
             max_3p = 10, max_5p = 10, endo = TRUE,
             adducts = c("H", "Na", "K", "NH4"), ppm_tol = 10)

wb_path <- build_workbook(spec, mets, opts = opts,
                          output_file = "demo_library.xlsx",
                          output_dir  = ".")
```

The Charge Envelopes sheet is the most computation-heavy (one isotope
pattern per metabolite per oxidation level); a live console progress
bar with an adaptive ETA tracks it. Sheets are assembled in memory and
written in bulk, so the Excel serialization itself takes seconds even
for libraries with tens of thousands of envelope rows.

## HTML/PDF report

`build_report()` renders a standalone report (requires `rmarkdown`)
with five summary plots -- charge envelope, isotope pattern,
truncation series, oxidation series, and fragment map. The individual
`plot_*()` functions are exported too, so any plot can be regenerated
or restyled on its own:

```{r report, eval = FALSE}
report <- build_report(spec, mets, opts = opts,
                       output_file = "demo_report",
                       output_format = "html",   # or "pdf"
                       plot_dir = ".")
```

## Orbitrap Exploris acquisition method export

Three targeted-acquisition tables are exported in the exact column
layout of the Orbitrap Exploris Method Editor's Targeted Mass filter
(Mass List Type "m/z & z", Time Mode "Start/End Time"):

```{r acquisition}
ms1 <- thermo_ms1_inclusion_list(mets[1:3], z_range = 5:8,
                                 max_oxid = 1, max_targets = 500)
head(ms1, 4)
```

* **`thermo_ms1_inclusion_list()`** -- every (metabolite, oxidation
  level, charge state) combination; import into a Full Scan
  experiment's Targeted Mass filter.
* **`thermo_ms2_prm_target_list()`** -- the same list narrowed to a
  smaller charge range and tagged with an HCD collision energy; import
  as the target list of a Targeted MS2 (PRM) scan. The default NCE
  (20%) is a starting point for PS oligonucleotide backbones, not a
  validated parameter -- optimize per method.
* **`ms2_fragment_reference()`** -- theoretical fragment ions per
  metabolite. This is *not* a Method Editor import (PRM records the
  full fragment spectrum); it is the lookup table for interpreting the
  acquired spectra afterward.

Both list functions cap output rows (`max_targets`) -- the Method
Editor accepts up to 150,000 rows, but PRM duty cycle degrades long
before that.

# Matching against acquired MS data

Point the pipeline at an mzML/mzXML file (or a two-column m/z-intensity
peak list) to annotate the library against real data:

```{r matching, eval = FALSE}
ms   <- parse_mzml("data.mzML")            # MS1 + MS2 spectra
feat <- extract_ms1_features(ms$ms1)
hits <- match_ms1(feat, mets, ppm_tol = 10,
                  z_range = 3:12, max_oxid = 6,
                  adducts = c("H", "Na", "K", "NH4"))
ann  <- annotate_metabolites(hits, ms, mets)   # adds MS2 confirmation
```

Matching proceeds in the order established by the eluforsen and FMVS
studies: (1) match each metabolite's theoretical m/z values across
charge states, adducts, and oxidation levels at the given ppm
tolerance; (2) score isotope-pattern fit (cosine similarity) and
charge-envelope consistency; (3) for hits with associated MS2 spectra,
confirm via McLuckey fragment matching and PS diagnostic ions.

Vendor raw files (Thermo `.raw`, Waters `.raw`, SCIEX `.wiff`) are
converted automatically when ProteoWizard `msconvert` is on the PATH.
mzML binary decoding uses a small Python helper when `python3` is
available and falls back to a pure-R decoder otherwise.

# The Shiny dashboard and CLI drivers

The repository root (not the installed package) carries three
ready-made entry points around the same engine:

* **`app.R`** -- a Shiny dashboard with sequence input, an editable
  custom-chemistry table, full parameter control, optional MS upload,
  a four-plot summary, and workbook/report/PRM/acquisition-list
  downloads. Launch with `shiny::runApp()` from a clone of the
  repository.
* **`run_custom_oligo.R`** -- the primary CLI template: copy it, edit
  the CONFIG block (sequence, overrides, parameters), and
  `Rscript run_custom_oligo.R`.

For a formula-engine regression check from the command line, run the
validation scripts in `tests/` (e.g.
`Rscript tests/test_mass_isotope.R`) or call `validate_ION337()`
directly from R.

# Performance notes

* The **Charge Envelopes** computation dominates runtime: one isotope
  pattern per metabolite per PS-oxidation level. Isotope patterns are
  memoized per formula (both the enviPat and built-in engines), so
  charge states and adducts add no repeated cost.
* To speed up a long run: disable endonuclease fragments, lower
  `max_oxid`, or narrow `z_range` -- each directly reduces the number
  of isotope-pattern calls.
* Workbook sheets are written in bulk (one write per sheet, styling
  restricted to headers and a few accent cells), so `saveWorkbook()`
  stays fast regardless of library size.
* `enviPat` prints a harmless note about the electron mass on some
  systems; it is cosmetic. Set `use_envipat = FALSE` to use the
  built-in engine instead.

# Caveats

* Dictionary entries flagged `verify = TRUE` (cEt, LNA, GalNAc3, dye
  and BPF terminal modifications) are best-estimate formulas --
  confirm against a known mass before quantitative use.
* Custom overrides are trusted as given; validate the parent mass
  against a reference when one exists.
* The `h_offset = 3.0046` convention exists only to reproduce the
  ION337 reference workbook; use 0 for standard work.
* The repository's `tests/` scripts are console-output validation
  scripts, not asserted unit tests.

# Session info {.unnumbered}

```{r sessioninfo}
sessionInfo()
```

# References {.unnumbered}

* Lippens J.L. *et al.* SynONIM: a Synthetic Oligonucleotide
  Nomenclature for Impurities and Modifications. *J Am Soc Mass
  Spectrom* (2024).
* Kim J. *et al.* Metabolite Profiling of the Antisense
  Oligonucleotide Eluforsen Using Liquid Chromatography-Mass
  Spectrometry. *Mol Ther Nucleic Acids* (2019).
* Liu R. *et al.* OligoDistiller: untargeted profiling of therapeutic
  oligonucleotide drugs and metabolites. *Anal Chem* (2025).
* Ye X. *et al.* Automatic metabolite identification of antisense
  oligonucleotides by full MS variable scanning. *J Chromatogr B*
  (2025).
* McLuckey S.A., Van Berkel G.J., Glish G.L. Tandem mass spectrometry
  of small, multiply charged oligonucleotides. *J Am Soc Mass
  Spectrom* (1992).
